---
title: 'Coursera: Practical Machine Learning Prediction Assignment'
author: "Benjamin Chan"
date: "Tuesday, October 07, 2014"
output:
  html_document:
    keep_md: yes
    toc: yes
---


> **Background**

> Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 


> **Data **

> The training data for this project are available here: 

> https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

> The test data are available here: 

> https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

> The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

> **What you should submit**

> The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

> 1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
> 2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 

> **Reproducibility **

> Due to security concerns with the exchange of R code, your code will not be run during the evaluation by your classmates. Please be sure that if they download the repo, they will be able to view the compiled HTML version of your analysis. 


# Get the training data

Read the training data into a data table.

```{r}
require(data.table)
setInternet2(TRUE)
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
D <- fread(url)
```

Make `classe` into a factor.

```{r}
classe <- factor(D[, classe])
is.factor(classe)
table(classe)
```

From [Velloso 2013, et al.](http://groupware.les.inf.puc-rio.br/public/papers/2013.Velloso.QAR-WLE.pdf):

> For feature extraction we used a sliding window approach
> with different lengths from 0.5 second to 2.5 seconds, with
> 0.5 second overlap. In each step of the sliding window approach
> we calculated features on the Euler angles (roll, pitch
> and yaw), as well as the raw accelerometer, gyroscope and
> magnetometer readings. For the Euler angles of each of the
> four sensors we calculated eight features: mean, variance,
> standard deviation, max, min, amplitude, kurtosis and skewness,
> generating in total 96 derived feature sets.

> In order to identify the most relevant features we used the
> feature selection algorithm based on correlation proposed by
> Hall [14]. The algorithm was configured to use a "Best First"
> strategy based on backtracking. 17 features were selected:
> in the belt, were selected the mean and variance of the roll,
> maximum, range and variance of the accelerometer vector,
> variance of the gyro and variance of the magnetometer. In
> the arm, the variance of the accelerometer vector and the
> maximum and minimum of the magnetometer were selected.
> In the dumbbell, the selected features were the maximum of
> the acceleration, variance of the gyro and maximum and
> minimum of the magnetometer, while in the glove, the sum
> of the pitch and the maximum and minimum of the gyro
> were selected.

```{r}
regexOther <- "classe|num_window|new_window|user_name"
regexBelt <- "^roll_belt$|^avg_roll_belt$|^var_roll_belt$|^accel_belt|^gyro_belt|^magnet_belt"
regexArm <- "^accel_arm|^magnet_arm"
regexDumbbell <- "^accel_dumbbell|^gyro_dumbbell|^magnet_dumbbell"
regexForearm <- "^pitch_forearm$|^gyro_forearm"
regex <- paste(regexOther, regexBelt, regexArm, regexDumbbell, regexForearm, sep="|")
varCol <- grep(regex, names(D))
DTrain <- D[, varCol, with=FALSE]
str(DTrain)
```

```{r}
DTrainAgg <- DTrain[,
                    list(roll_beltAvg = mean(roll_belt, na.rm=TRUE),
                         roll_beltVar = var(roll_belt, na.rm=TRUE)),
                    list(user_name, num_window, classe)]
DTrainAgg <- merge(DTrainAgg,
                   DTrain[new_window == "yes", list(user_name, num_window, classe, avg_roll_belt, var_roll_belt)],
                   by=c("user_name", "num_window", "classe"))
```

Subset columns with names beginning with *roll*, *pitch*, *yaw*, *total_accel*, *gyros*, *accel*, *magnet*
These will be the predictor variables.
There are a lot of missing values in the other variables.

```{r, eval=FALSE}
regex <- "^roll|^pitch|^yaw|^total_accel|^gyros|^accel|^magnet"
predVarCol <- grep(regex, names(D))
predVar <- D[, predVarCol, with=FALSE]
DTrain <- data.table(classe, predVar)
str(DTrain)
sum(is.na(DTrain))
```

Preprocess the prediction variables by centering and scaling.

```{r}
require(caret)
preProc <- preProcess(predVar)
preProc
DTrainCS <- data.table(classe, predict(preProc, predVar))
```

Check for near zero variance.

```{r}
nearZeroVar(DTrainCS, saveMetrics=TRUE)
```

Examine groups of prediction variables.

```{r}
histGroup <- function (data, regex) {
  col <- grep(regex, names(data))
  col <- c(col, which(names(data) == "classe"))
  require(reshape2)
  n <- nrow(data)
  DMelted <- melt(data[, col, with=FALSE][, rownum := seq(1, n)], id.vars=c("rownum", "classe"))
  require(ggplot2)
  ggplot(DMelted, aes(x=classe, y=value)) +
    geom_violin(aes(color=classe, fill=classe), alpha=1/2) +
#     geom_smooth(aes(group=1), method="glm") +
    facet_wrap(~ variable, scale="free_y") +
    scale_color_brewer(palette="Spectral") +
    scale_fill_brewer(palette="Spectral") +
    labs(x="", y="") +
    theme(legend.position="none")
}
histGroup(DTrainCS, "belt")
histGroup(DTrainCS, "[^(fore)]arm")
histGroup(DTrainCS, "dumbbell")
histGroup(DTrainCS, "forearm")
```


# Build a prediction model

Set up the parallel clusters.

```{r, eval=FALSE}
require(parallel)
require(doParallel)
cl <- makeCluster(2)
registerDoParallel(cl)
```

Set the control parameters.

```{r, eval=FALSE}
ctrl <- trainControl(classProbs=TRUE,
                     savePredictions=TRUE,
                     allowParallel=TRUE)
```

Preprocess with PCA.

```{r, eval=FALSE}
preProc <- preProcess(predVar, method="pca")
trainPC <- predict(preProc, predVar)
# M1 <- train(classe ~ ., data=trainPC, method="glm")
# testPC <- predict(preProc, testing[predVar])
# hat1 <- predict(M1, testPC)
# confusionMatrix(testing$diagnosis, hat1)
```

Fit model over the tuning parameters.

```{r, eval=FALSE}
method <- "C5.0"
M0 <- train(classe ~ ., data=DTrainCS, method=method, trControl=ctrl)
M1 <- train(classe ~ ., data=trainPC, method=method, trControl=ctrl)
```

Stop the clusters.

```{r, eval=FALSE}
stopCluster(cl)
```

Evaluate the model.

```{r, eval=FALSE}
M0
confusionMatrix(predict(M0), classe)
M1
confusionMatrix(predict(M1), classe)
```
